\thetitle

### 2021-05-01 writing/reviews/icfp21-rebuttal.md

Thanks, we will consider these alternative names for the final paper.

> My only issue at this point is that the reader is left with numnbers only,
> while I think a reader would like to see selected examples, for example about
> which blame labels have been provided by Natural and which by Transient, so to
> see that one approach took a longer path, as well as similar examples the
> reader can learn from. Numbers do not seem to teach the whole story in this
> part of the paper.

This is a good idea, thanks. In the final paper we will try to incorporate one
or more examples that illustrate the difference in trails between the three
mechanisms.

Review #45D [R1]
===========================================================================

> First, the performance figures cited from [38] (worst case 5.4x
> slowdown) are used incorrectly: those numbers are about transient
> /without/ blame (see Section 6.2 of [38]), and given that "with the
> blame map turned off, the Transient semantics also runs these programs
> in a short amount of time and well within the memory limits" there
> appears to be no contradiction here. When blame /is/ enabled, [38]
> reports a worst-case 18x slowdown, which may be more in line with the
> results reported in this work, and as a result it's not actually clear
> that the earlier results are skewed in a way that demands explanation.

Yes you are right, this is our mistake: we meant to use the number with blame
here. After looking at the plot in Section 6.2 of [38] we believe the right
numbers are 6.2x and 17.2x (the 18x is an upper bound).

These numbers are still orders-of-magnitude better than Shallow. We believe this
gap needs an explanation.

> Claim (2), that local variables in Reticulated Python are dynamically
> typed because no types are ascribed to them, is false; Reticulated
> Python uses a local type inference system to give types to local
> variables (see Section 3.1.3 of [36]).

Correct, Reticulated uses local type inference. We need to acknowledge this
fact. Our overall claim stands, however, because Reticulated's inference often
picks the dynamic type for local variables.

This behavior came as a surprise to us too, and we had to manually inspect the
output of `retic --mgd-transient --print file.py` for several benchmarks to
convince ourselves.

Section 5.4.4 of Greenman's dissertation presents a tiny example where
Reticulated infers `Dyn` for a local variable.

> Claim (4) that "on large programs, Reticulated suffers from high
> overhead" is unsubstantiated here or in cited prior work, and is also
> unclear -- overhead compared to what? The text compares it (vaguely) to
> Typed Racket (without specifying which semantics), but this isn't a
> useful comparison given the different baseline performance of Racket and
> CPython. It's not clear if the point is that performance of this
> benchmark in blame-free Reticulated is seriously slow compared to
> regular Python, or if the program itself is just slow in Python (in
> which case, what's the relevance to this paper?).

You are right, the prose here does not explain the comparison here well.
Especially, Typed Racket has no place in this bullet.

First off, our goal was to run Reticulated on one large benchmark to see if it
ever had a huge "timeout-level" slowdown.

To this end, we converted our fully-typed `sieve` benchmark from Racket to
Reticulated. Then we ran this fully-typed Reticulated program twice: without
blame and with blame. Adding blame gave a huge slowdown.

This little experiment suggests that Reticulated does not have any crucial
implementation techniques that Shallow Racket is missing.

> The characterization of the Monotonic semantics for references [22] as a
> variant of natural isn't correct--its exception-raising and blame
> behavior for references (though not for functions) is quite different
> from that of Natural. For example, a monotonic reference imported from
> untyped code into two incompatible typed contexts will error
> immediately, before being used.

Indeed this characterization is not right, monotonic is more than a mere variant
of Natural. We will fix this description in the final paper.

### 2021-05-02 writing/reviews/icfp21-rebuttal.md

Review #45D [R1]
===========================================================================

> First, the performance figures cited from [38] (worst case 5.4x
> slowdown) are used incorrectly: those numbers are about transient
> /without/ blame (see Section 6.2 of [38]), and given that "with the
> blame map turned off, the Transient semantics also runs these programs
> in a short amount of time and well within the memory limits" there
> appears to be no contradiction here. When blame /is/ enabled, [38]
> reports a worst-case 18x slowdown, which may be more in line with the
> results reported in this work, and as a result it's not actually clear
> that the earlier results are skewed in a way that demands explanation.

Yes you are right, this is our mistake: we meant to use the number with blame
here. After looking at the plot in Section 6.2 of [38] we believe the right
numbers are 6.2x and 17.2x (the 18x is an upper bound).

These numbers, however, are still orders-of-magnitude better than Shallow. Hence, 
we believe this gap needs an explanation.

@@

> Claim (2), that local variables in Reticulated Python are dynamically
> typed because no types are ascribed to them, is false; Reticulated
> Python uses a local type inference system to give types to local
> variables (see Section 3.1.3 of [36]).

You are right and we should had been more precise here. Reticulated does attempt to 
infer types for local variables. However, our experience indicates Reticulated's inference often
picks the dynamic type for local variables.

@@

This behavior came as a surprise to us too, and we had to manually inspect the
output of `retic --mgd-transient --print file.py` for several benchmarks to
convince ourselves.

Section 5.4.4 of Greenman's dissertation presents a tiny example that
explains how Reticulated can infer `Dyn` for a local variable.

chrdimo: is it worth it expanding on this here?

@@

> Claim (4) that "on large programs, Reticulated suffers from high
> overhead" is unsubstantiated here or in cited prior work, and is also
> unclear -- overhead compared to what? The text compares it (vaguely) to
> Typed Racket (without specifying which semantics), but this isn't a
> useful comparison given the different baseline performance of Racket and
> CPython. It's not clear if the point is that performance of this
> benchmark in blame-free Reticulated is seriously slow compared to
> regular Python, or if the program itself is just slow in Python (in
> which case, what's the relevance to this paper?).

Thank for pointing this out. The prose here does not explain clearly our
evidence for this claim.

First off, the overall goal was to run Reticulated (with and without blame) on one 
large benchmark to see if it ever had a huge "timeout-level" slowdown,
similar to the ones we observe in our ``rational programmer'' experiment.

To this end, we converted our fully-typed `sieve` benchmark from TR to
Reticulated. Then we ran this fully-typed Reticulated program twice: without
blame and with blame. Adding blame gave a huge slowdown.

This little experiment suggests that Reticulated does not employ any crucial
implementation techniques that can mitigate the slowdown we observe with Shallow Racket.

chrdimo: Is the translation faithful in terms of types?

chrdimo: I think there is a valid point here that the claim comes out as
too strong but what we have a single data point. I think we should also
admit this and say that we will tone down the claim.

@@

> The characterization of the Monotonic semantics for references [22] as a
> variant of natural isn't correct--its exception-raising and blame
> behavior for references (though not for functions) is quite different
> from that of Natural. For example, a monotonic reference imported from
> untyped code into two incompatible typed contexts will error
> immediately, before being used.

Indeed this characterization is not right, when it comes to boxes
monotonic is more than a mere variant of Natural. We will fix this
description in the final paper.

chrdimo: Is monotonic implemented? What happens with hashes, objects and
other containers?

### 2021-05-03 writing/reviews/icfp21-rebuttal.md

Yes you are right, this is our mistake: we meant to use the number with blame
here. After looking at the plot in Section 6.2 of [38] we believe the right
numbers are 6.2x and 17.2x (the 18x is an upper bound).

These numbers, however, are still orders-of-magnitude better than Shallow. Hence, 
we believe this gap needs an explanation.

> Claim (2), that local variables in Reticulated Python are dynamically
> typed because no types are ascribed to them, is false; Reticulated
> Python uses a local type inference system to give types to local
> variables (see Section 3.1.3 of [36]).

You are right and we should have been more precise here. Reticulated
does attempt to infer types for local variables. However, our
experience indicates Reticulated's inference often picks type Dynamic
for local variables. Section 5.4.4 of Greenman's dissertation presents
a tiny example that explains how Reticulated can infer `Dyn` for a
local variable.

This behavior came as a surprise to us too. We manually inspected the
output of `retic --mgd-transient --print file.py` for several benchmarks to
convince ourselves.

@@

> Claim (4) that "on large programs, Reticulated suffers from high
> overhead" is unsubstantiated here or in cited prior work, and is also
> unclear -- overhead compared to what? The text compares it (vaguely) to
> Typed Racket (without specifying which semantics), but this isn't a
> useful comparison given the different baseline performance of Racket and
> CPython. It's not clear if the point is that performance of this
> benchmark in blame-free Reticulated is seriously slow compared to
> regular Python, or if the program itself is just slow in Python (in
> which case, what's the relevance to this paper?).

Thank for pointing this out. The prose here does not explain clearly our
evidence for this claim.

The goal was to run Reticulated (with and without blame) on one
reasonably large benchmark to see if it ever had a huge
"timeout-level" slowdown, similar to the ones we observe in our
``rational programmer'' experiment.

To this end, we converted our fully-typed `sieve` benchmark from TR to
Reticulated. Then we ran this fully-typed Reticulated program twice:
with and without blame. The `with blame` variant experienced a huge
slowdown.

This little experiment suggests that Reticulated does not employ any
crucial implementation techniques that can mitigate the slowdown we
observe with Shallow Racket. Of course, all we have is a single data
point, and this must come thru more clearly in our prose. We would
hope that this submission and the parallel submission on Shallow
Racket's construction inspires the Reticulaed team to report data on
larger benchmarks than the ones in their published papers.

### 2021-05-03 writing/reviews/icfp21-rebuttal.md

Thank you for pointing this out. The prose here does not explain clearly our
evidence for this claim.

The goal was to run Reticulated (with and without blame) on one
reasonably large benchmark to see if it ever had a huge
"timeout-level" slowdown, similar to the ones we observe in our
``rational programmer'' experiment.

To this end, we converted our fully-typed `sieve` benchmark from TR to
Reticulated. Then we ran this fully-typed Reticulated program twice:
with and without blame. The `with blame` variant experienced a huge
slowdown.

This little experiment suggests that Reticulated does not employ any
crucial implementation techniques that can mitigate the slowdown we
observe with Shallow Racket. Of course, all we have is a single data
point, and this must come thru more clearly in our prose. We would
hope that this submission and the parallel submission on Shallow
Racket's construction inspires the Reticulated team to report data on
larger benchmarks than the ones in their published papers.

### 2021-05-03 writing/reviews/icfp21-rebuttal.md

Thank you for pointing this out. The prose here does not explain clearly our
evidence for this claim.

The goal was to run Reticulated (with and without blame) on one
reasonably large benchmark to see if it ever had a huge
"timeout-level" slowdown, similar to the ones we observe in our
``rational programmer'' experiment.

To this end, we converted our fully-typed `sieve` benchmark from TR to
Reticulated. Then we ran this fully-typed Reticulated program twice:
with and without blame. The `with blame` variant experienced a huge
slowdown.

This little experiment suggests that Reticulated does not employ any
crucial implementation techniques that can mitigate the slowdown we
observe with Shallow Racket. Of course, all we have is a single data
point, and this must come through more clearly in our prose.

We hope that this submission and the parallel submission on Shallow
Racket's construction will inspire further investigation of Transient as a
viable production-level approach.

### 2021-05-04 writing/landscape.tex
Remove the clever table layout in lieu of a sentence description

The following table summarizes the illustration:
%% -----------------------------------------------------------------------------
\begin{center}
  \begin{tabular}{l|ll@{}}
                        &        \multicolumn{2}{|c}{the evaluation of} \\
\relax                  & {\tt public-data}               & {\tt secret-data} in {\tt client}, \\
{under the $\cdot$ semantics yields} & & \\ \hline
%% ----------------------------------------------------------------------------------------
\qquad {\it Natural\/}   & error, blaming {\tt pack-lib}  & error, blaming {\tt pack-lib} \\
     		   	 &                                & \\
\qquad {\it Transient\/} & no error   		          & error, blaming the boundary crossings \\
      		  	 &    			          & \quad {\tt pack-lib} to {\tt typed-pack-lib} \\
               &                     & \quad {\tt pack-lib} to {\tt crypto-pack-lib} \\
\qquad {\it Erasure\/}   & no error 		          & no error \\
       	    		 &   \multicolumn{2}{|c}{but {\em Erasure} does signal an error on list access}
\end{tabular}
\end{center}
%% -----------------------------------------------------------------------------

>>>>

The following table summarizes the illustration. Each cell describes the result
of evaluating the column's definition (in {\tt client}) under the row's
semantics.
%% -----------------------------------------------------------------------------
\begin{center}
  \begin{tabular}{l|ll}
                  & {\tt public-data}              & {\tt secret-data}                                 \\
\hline %% ----------------------------------------------------------------------------------------
{\it Natural\/}   & error, blaming {\tt pack-lib}  & error, blaming {\tt pack-lib}                     \\
                  &                                &                                                   \\
{\it Transient\/} & no error                       & error, blaming the boundary crossings             \\
                  &                                & \quad {\tt pack-lib} to {\tt typed-pack-lib}, and \\
                  &                                & \quad {\tt pack-lib} to {\tt crypto-pack-lib}     \\
                  &                                &                                                   \\
{\it Erasure\/}   & no error*                      & no error*                                         \\
                  &   \multicolumn{2}{|c}{*but {\em Erasure} does signal an error on list access}
\end{tabular}
\end{center}
%% -----------------------------------------------------------------------------
